---
title: "Week 2 Assignment"
author: "Arghavan Kassraie"
output: html_document
---

```{r setup, include=FALSE}
suppressPackageStartupMessages({
  library(AnnotationHub)     # to fetch genomes, annotations, etc.
  library(ensembldb)         # to manipulate ensembl gene annotations
  library(GenomicRanges)     # data structure for genomic ranges 
  library(epiwraps)          # visualization
})
```

## 1. Mouse EnsDb Object (GRCm38)


```{r query_mouse_ensdb}
ah <- AnnotationHub()
query_result_1 <- query(ah, c("EnsDb", "Mus musculus", "102", "GRCm38"))
query_result_1
```

```{r load_mouse_ensdb}
mouse_ensdb <- ah[["AH89211"]]
mouse_ensdb
```

## 2. Mouse Genome Sequence


```{r query_mouse_genome}
query_result_2 <- query(ah, c("dna_sm", "Mus musculus", "GRCm38"))
query_result_2
```

```{r load_mouse_genome}
mouse_gs_2bit <- ah[["AH49775"]]
mouse_gs_2bit
```

## 3. Drosophila Genome Sequence


```{r query_drosophila_genome}
query_result_3 <- query(ah, c("dna_sm", "Drosophila", "BDGP6"))
query_result_3 
```

```{r load_drosophila_genome}
dros_gs_2bit <- ah[["AH49674"]]
dros_gs_2bit
```

## 4. Protein-coding Genes Analysis

Count the number of unique Ensembl gene IDs and gene symbols for protein-coding genes.

```{r protein_coding_genes}
# Extract protein-coding genes
pc_genes <- genes(mouse_ensdb,
                  filter = GeneBiotypeFilter("protein_coding"))

# Count unique gene IDs and symbols
n_gene_ids <- length(unique(pc_genes$gene_id))
n_gene_symbols <- length(unique(pc_genes$gene_name))

cat("Number of unique Ensembl gene IDs:", n_gene_ids, "\n")
cat("Number of unique gene symbols:", n_gene_symbols, "\n")
```

## 5. Distribution of Spliced Transcript Lengths

Analyze the distribution of spliced lengths for protein-coding transcripts.

```{r transcript_lengths}
# Extract protein-coding exons and group by transcript
exons_by_tx <- exonsBy(mouse_ensdb, by = "tx",
                       filter = TxBiotypeFilter("protein_coding"))

# Compute spliced transcript lengths per transcript
tx_lengths <- sapply(exons_by_tx, function(exon_gr) sum(width(exon_gr)))
```

```{r plot_transcript_distribution, fig.width=8, fig.height=6}
# Plot distribution of spliced transcript lengths
hist(tx_lengths,
     breaks = 100,
     main = "Distribution of Spliced Lengths\nProtein-coding Transcripts",
     xlab = "Transcript Length",
     ylab = "Frequency",
     col = "blue",
     border = "white")
```


It can be seen that most transcripts have a small length, with the distribution heavily skewed towards shorter transcripts. 
