---
title: "Week 7 assignment"
author: "Arghavan Kassraie"
date: "`r Sys.Date()`"
output: html_document
---

```{r }
suppressPackageStartupMessages({
  library(epiwraps)
  library(AnnotationHub)
  library(MotifDb)
  library(universalmotif)
  library(ensembldb)
  library(ggplot2)
  library(motifmatchr)
  library(sechm)
  library(BSgenome.Mmusculus.UCSC.mm10)
  library(limma)
  library(betterChromVAR)
})
```

# Part 1

## Download the ATAC-seq bam files

```{r}
if (!file.exists("atac.chr19.bam")) {
  download.file("https://ethz-ins.org/content/w7/atac.chr19.bam",
                "atac.chr19.bam", mode = "wb")
  download.file("https://ethz-ins.org/content/w7/atac.chr19.bam.bai",
                "atac.chr19.bam.bai", mode = "wb")
}
```

## Preparing tracks 

```{r }
bam <- "atac.chr19.bam"

bam2bw(bam, output_bw = "NF_cuts.bw", paired = TRUE, binWidth = 1L,
       type = "ends", minFragLength = 30, maxFragLength = 120,
       trim = 4L, extend = 3L, forceSeqlevelsStyle = "Ensembl")

bam2bw(bam, output_bw = "mono_centers.bw", paired = TRUE, binWidth = 5L,
       type = "center", minFragLength = 140, maxFragLength = 220,
       trim = 4L, extend = 10L, forceSeqlevelsStyle = "Ensembl")
```

## Get the sites with high confidence motif occurrences on chr19

```{r}
ah <- AnnotationHub()
genome <- ah[["AH68356"]]          # mouse genome (2bit)
chr19  <- import(genome)["19"] #get the sequence for 

# get CTCF motif 
motif  <- MotifDb::query(MotifDb, c("CTCF", "Mus"))[[1]]
motif2 <- convert_motifs(motif, class = "TFBSTools-PFMatrix")

# find motic matches 
moi <- motifmatchr::matchMotifs(motif2, chr19, out = "positions",
                                p.cutoff = 1e-5)[[1]]
#convert to GRanges
moi <- as(setNames(moi, names(chr19)), "GRanges")
length(moi)
```

## Plot both signals around the motif occurrences

```{r}
tracks <- c(NF_cuts = "NF_cuts.bw", mono_centers = "mono_centers.bw")

sm <- signal2Matrix(tracks, moi, w = 5, extend = 300)

plotEnrichedHeatmaps(sm, trim = 0.95, multiScale = TRUE,
                     axis_name = "CTCF motif", use_raster = FALSE)
```

To make the relationship between the two signals clearer, we can overlay their aggregated profiles:

```{r}
d <- meltSignals(sm)
ggplot(d, aes(position, mean, colour = sample)) +
  geom_line(linewidth = 1.2) +
  labs(x = "Position relative to CTCF motif (bp)",
       y = "Mean signal",
       title = "ATAC-seq signal around CTCF motifs on chr19") +
  theme_bw()
```


# Part 2

## Load the data

```{r}
if (!file.exists("peakCounts.rds")) {
  download.file("https://ethz-ins.org/content/mouse_mm38_hippocampus.peakCounts.SE.rds",
                "peakCounts.rds", mode = "wb")
}
se <- readRDS("peakCounts.rds")
se
colData(se)
```

## Annotate regions 

```{r}
se <- GenomeInfoDb::keepStandardChromosomes(se, pruning.mode = "coarse")
seqlevelsStyle(se) <- "UCSC"
se <- chromVAR::addGCBias(se, genome = BSgenome.Mmusculus.UCSC.mm10)
```

## Get motifs and scan regions

```{r}
motifs <- query(MotifDb, "HOCOMOCOv13")
motifs <- do.call(TFBSTools::PWMatrixList,
                  convert_motifs(motifs, class = "TFBSTools-PWMatrix"))
names(motifs) <- make.unique(sapply(motifs, \(x) gsub("\\..+", "", x@name)))

matches <- motifmatchr::matchMotifs(motifs, se,
                                    genome = BSgenome.Mmusculus.UCSC.mm10)
```

## Get motif deviations

```{r}
dev <- betterChromVAR(se, matches)
dev
```

## Differential analysis

```{r}
# infer sex and condition from sample names
dev$sex       <- ifelse(grepl("female", colnames(dev), ignore.case = TRUE),
                        "female", "male")
dev$condition <- ifelse(grepl("FSS", colnames(dev)), "Stressed", "Control")

table(dev$sex, dev$condition)

design <- model.matrix(~condition + sex, data = as.data.frame(colData(dev)))
fit <- eBayes(lmFit(assay(dev, "z"), design))
res <- as.data.frame(topTable(fit, coef = "sexmale", number = Inf))
res$motif <- rownames(res)

head(res, 10)
```

```{r}
ggplot(res, aes(logFC, -log10(adj.P.Val), label = motif)) +
  geom_text(size = 3) +
  labs(x = "log fold-change (male vs. female)",
       y = "-log10(adjusted p-value)",
       title = "Sex-differential motif accessibility") +
  theme_bw()
```

```{r}
top_motifs <- head(rownames(res), 10)
sechm(dev, top_motifs,
      top_annotation = c("sex", "condition"),
      gaps_at = "sex",
      assayName = "z",
      use_raster = FALSE)
```
