---
title: "Week 9 assignment"
author: "Arghavan Kassraie"
output: html_document
---


```{r libraries}
suppressPackageStartupMessages({
  library(GenomicRanges)
  library(epiwraps)
  library(ggplot2)
  library(rGREAT)
  library(rtracklayer)
  library(ComplexHeatmap)
})
```

## Download data
```{r download}
options(timeout = 6000)
download.file(
  "https://ethz-ins.org/content/clustering_assignment.zip",
  "clustering_assignment.zip"
)
unzip("clustering_assignment.zip")
```

## Prepare regions and tracks
```{r regions}
tracks <- c(
  "Creb1.bw",
  "Creb3.bw",
  "Creb3L1.bw"
)
peak_files <- c(
  "Creb1.bed",
  "Creb3.bed",
  "Creb3L1.bed"
)
# Import peaks
peaks <- lapply(peak_files, rtracklayer::import.bed)
# Keep high confidence peaks
peaks <- lapply(peaks, function(x) x[x$score > 800])
# Merge regions
regions <- reduce(unlist(GRangesList(peaks)))
```

## Clustering and visualisation
```{r clustering, fig.width=10, fig.height=8}
ese <- signal2Matrix(tracks, regions, extend = 2000)

set.seed(123)
cl <- clusterSignalMatrices(ese, k = 4)
rowData(ese)$cluster <- cl

mycolors <- c("1"="red", "2"="blue", "3"="darkgreen", "4"="black")

plotEnrichedHeatmaps(ese,
                     row_split  = rowData(ese)$cluster,
                     mean_color = mycolors,
                     colors     = c("white", "darkred"),
                     use_raster = FALSE)
```

## GO enrichment on cluster 3
```{r enrichment, fig.width=9, fig.height=6}
split_regions <- split(
  rowRanges(ese),
  rowData(ese)$cluster
)
res <- great(
  split_regions[["3"]],
  gene_sets  = "GO:BP",
  tss_source = "hg38",
  background = regions,
  cores      = 2
)
bp <- getEnrichmentTables(res)
ggplot(
  head(bp, 15),
  aes(
    fold_enrichment,
    reorder(description, p_adjust),
    size  = observed_region_hits,
    color = -log10(p_adjust)
  )
) +
  geom_point() +
  scale_color_viridis_c() +
  labs(
    title = "GO Enrichment Analysis – Cluster 3",
    x = "Fold enrichment",
    y = NULL
  ) +
  theme_bw()
```
Cluster 1 is primarily driven by Creb1-specific binding, whereas Cluster 3 shows binding mainly exclusive to Creb3. In contrast, Cluster 2 highlights shared binding sites strongly occupied by all three factors, and Cluster 4 represents regions co bounded by Creb1 and Creb3L1 with minimal Creb3 involvement. GO enrichment analysis of cluster 3 reveals over representation of terms related to cellular response to stimulus, cell communication, and cell surface receptor signaling, consistent with Creb3's role as a stress responsive transcription factor. Together, these results show that while CREB-family members share many binding sites, Creb3 and Creb3L1 each occupy distinct binding profiles reflecting their specific biological roles.

